---
title: "Healthcare cost analysis (Medicaid)"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    social: menu
    source: embed
---

```{r setup, include=FALSE}
library(flexdashboard)
library(readr)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggvis)
library(tidyr)
library(broom)
library(ggpubr)
library(plotly)
library(devtools)
# Restore the object
data <- readRDS(file = "df_13_17_pop.rds")
# removing data with State == XX
data_A <- data %>% filter(State != "XX")
#changing Product_Name as Product
names(data_A)[names(data_A) == "Product Name"] <- "Product"
# summarizing the data
data_B <- data_A %>% group_by(`State`, `Year`,`Quarter` ) %>% 
  summarize(tot_amt_reim = sum(`Total Amount Reimbursed`),
    tot_unt_reim = sum(`Units Reimbursed`),tot_nm_pres = sum(`Number of Prescriptions`),
    pop = mean(`Population`, na.rm = TRUE)) %>% mutate(unit_cost = tot_amt_reim/tot_unt_reim,
                      amt_capita = tot_amt_reim/pop)
data_B_TN <- data_A %>% filter(State == "TN") %>% 
  group_by(`Year`,`Quarter` )%>% 
  summarize(tot_amt_reim = sum(`Total Amount Reimbursed`),
            tot_unt_reim = sum(`Units Reimbursed`),tot_nm_pres = sum(`Number of Prescriptions`),
            pop = mean(`Population`, na.rm = TRUE)) %>%
  mutate(unit_cost = tot_amt_reim/tot_unt_reim,amt_capita = tot_amt_reim/pop)

```

### **Total cost of Medicaid over the years for the whole nation and TN specifically**

```{r, fig.height=8, fig.width=9.5}
# Q1a: Total cost over the years for country
Yearly_cost <- data_B %>% group_by(`Year`) %>% summarize(total_cost = sum(`tot_amt_reim`)) 
# Q1a: Total cost over the years for TN
Yearly_cost_TN <- data_B_TN %>% group_by(`Year`) %>% summarize(total_cost = sum(`tot_amt_reim`))

blue.bold.italic.16.text <- element_text(face = "bold.italic", color = "blue", size = 16)
black.bold.italic.16.text <- element_text(face = "bold.italic", color = "black", size = 16)
black.bold.italic.12.text <- element_text(face = "bold.italic", color = "black", size = 12)
black.bold.italic.6.text <- element_text(face = "bold.italic", color = "black", size = 6)
gray.bold.italic.16.text <- element_text(face = "bold.italic", color = "gray4", size = 16)
red.bold.italic.16.text <- element_text(face = "bold.italic", color = "red4", size = 16)
# bar plot showing the increase in the cost
par(mfrow = c(1,2))
# For whole country
ggplot(Yearly_cost, aes(reorder( Year, total_cost), total_cost)) + geom_col(width = 0.6, fill = "red4") +
labs(title = "Medicaid cost over the years", x = "Year",
y = "Total cost (billions)") + theme(title = black.bold.italic.16.text,
axis.title = black.bold.italic.16.text, axis.text = element_text(size = 8, face = 'bold')) + 
scale_y_continuous(name = "Total cost (billions)",
breaks = c(10000000000,20000000000,30000000000,
40000000000,50000000000,60000000000 ),
labels = c(10,20,30,40,50, 60))
# For TN
ggplot(Yearly_cost_TN, aes( Year, total_cost)) +
geom_col(width=0.6, fill = "navy") +
labs(title = "Medicaid cost over the years in TN", x = "Year",
y = "Total cost (billions)") +   theme(title = black.bold.italic.16.text, axis.title = black.bold.italic.16.text,
                                      axis.text = element_text(size = 8, face = 'bold') ) +
scale_y_continuous(name = "Total cost (billions)",
breaks = c(100000000,200000000,400000000,600000000,
800000000,1000000000,1200000000 ),
labels = c(0.1,0.2,0.4,0.6,0.8,1.0,1.2))
```

### **Top ten states spending most over the years**

```{r, fig.height=8, fig.width=9.5}
#Q2 State wise total cost over the years
top_states <- data_B %>% group_by(`State`) %>%
  summarize(total_cost = sum(`tot_amt_reim`)) %>% 
  arrange(desc(total_cost))
#Q2a States with highest per capita expenditure
top_st_per_cap <- data_B %>% group_by(`State`) %>% summarize(total_cost_per_cap = sum(`amt_capita`)) %>% 
  arrange(desc(total_cost_per_cap)) 

# plotting
par(mfrow = c(1,2))
head(top_states, 10) %>% ggplot(aes(reorder( State, total_cost),total_cost)) +
  geom_col(width=0.5, fill = "red4") +
  labs(title = "States with highest cost in last five years",
       x = "State", y = "Total cost (billions)") +
  theme(title = black.bold.italic.16.text,
        axis.title = black.bold.italic.16.text) + 
  scale_y_continuous(name = "Total cost (billions)",
  breaks = c(5000000000,10000000000,20000000000,
  30000000000,40000000000),
  labels = c(0.5,10,20,30, 40))
# States with highest per capita expenditure
head(top_st_per_cap, 10) %>% ggplot(aes(reorder( State, total_cost_per_cap),total_cost_per_cap)) +
geom_col(width=0.5, fill = "navy") + 
labs(title = "States with highest cost per person",
x = "State", y = "Cost per person ($)") +
theme(title = black.bold.italic.16.text, axis.title = black.bold.italic.16.text)
```

### **States spending less over the years**

```{r, fig.height=8, fig.width=9.5}
# Q3a: State wise total cost over the years(Bottom one) 
tail(top_states, 10) %>% ggplot(aes(reorder( State, total_cost),total_cost)) +
geom_col(width= 0.5, fill = "darkgreen") +
labs(title = "States with lowest cost in last five years",
x = "State", y = "Total cost (billions)") +
theme(title = black.bold.italic.16.text,
axis.title = black.bold.italic.16.text) + 
scale_y_continuous(name = "Total cost (billions)",
breaks = c( 100000000,200000000,400000000,600000000,
                800000000,1000000000 ),
labels = c(0.1 ,0.2,0.4,0.6,0.8,1.0))

# Lowest per capita
tail(top_st_per_cap, 10) %>% ggplot(aes(reorder( State, total_cost_per_cap),total_cost_per_cap)) +
geom_col(width = 0.5, fill = "gold4") +
labs(title = "States with lowest cost per person", x = "State",
y = "Cost per person ($)") +
theme(title = black.bold.italic.16.text,
axis.title = black.bold.italic.16.text)
```


### **States grouping (Hierarchical clustering)**

```{r, fig.height=10, fig.width=16}
# Q3a: State wise total cost over the years(Bottom one) 
states <- data_B %>% group_by(`State`) %>%
  summarize(total_cost = sum(`tot_amt_reim`),
  total_cost_per_cap = sum(`amt_capita`)) %>% 
  arrange(desc(total_cost_per_cap)) 

# normalization removing 1st chr variable
z = states[, -1]
# calculating mean and sd of each column
m <- apply(z, 2, mean)
s <- apply(z, 2, sd)
z <- scale(z, m, s)
# calculating eucleaden dist
distance = dist(z)
# cluster dendo with complete linkage
hc_c <- hclust(distance)
plot(hc_c, labels = states$State, hang = -1, cex = 1.5,
     main = "Landscape of States based on expenditure ",
     xlab = "States",
     ylab = "Height", cex.lab = 1.6, cex.axis = 1.8, cex.main = 2.8, font.main=4, font.lab=4) 

```


### **States grouping (k-means clustering)**

```{r, fig.height=10, fig.width=18}
# k means clustering with normalized values
kc <- kmeans(z, 5)
plot(total_cost_per_cap ~ total_cost, states, col = kc$cluster,
     pch = 19,cex = 1.5,
     main = "Distribution of States",
     xlab = "Total cost (billions)",
     ylab = "Cost/person", ylim = c(200,1800), xaxt = 'n',
     cex.lab = 1.5, cex.axis = 1.8, cex.main = 2.8, font.main=4, font.lab=4)
with(states, text(total_cost_per_cap ~ total_cost,
                  labels = State, pos = 4, cex = .9)) 
axis(side = 1, at = c(150000000,5000000000,10000000000,
                 15000000000, 25000000000, 40000000000),
     labels = c("0.15","5.0","10","15", "25","40"))

     

```




### **Top five drugs costing the Medicaid**

```{r, fig.height=8, fig.width=9.5}
# Q4: Top drugs 
data_Drg <- data_A %>% group_by(`Product`) %>% 
summarize(tot_amt_reim = sum(`Total Amount Reimbursed`),
tot_unt_reim = sum(`Units Reimbursed`),
tot_nm_pres = sum(`Number of Prescriptions`),
pop = mean(`Population`, na.rm = TRUE)) %>% 
mutate(unit_cost = tot_amt_reim/tot_unt_reim) %>% 
arrange(desc(tot_amt_reim))

top10drugs <- head(data_Drg, 10)
top5drugs <- head(data_Drg, 5)

dis_prod <- data.frame('Product'=c('HUMIRA - A','METHYLPHEN','INVEGA SUS','TRUVADA', 'HARVONI','ABILIFY', 'ADDERALL X','ADVAIR DIS','SOVALDI 40', 'FLOVENT HF','VYVANSE', 'PROVENTIL','LANTUS',
'FOCALIN XR','HUMALOG','LYRICA','SEROQUEL X'),
'Dis'= c('Arthritis','ADHD','Schizophrenia', 'AIDS','HCV', 'Schizophrenia', 'ADHD','Asthma', 'HCV', 'Asthma', 'BED', 'Asthma', 'Diabetes','ADHD','Diabetes', 'Fibromyalgia','Schizophrenia'),
'gen_cost' = c(1789,0.21,1300,7,1094,1,3,5,1000,10,16,5,20,1,22,5,6))

top10drugs_dis <- merge(top10drugs,dis_prod, by = "Product" )
top5drugs_dis <- merge(top5drugs,dis_prod, by = "Product" )
# plotting
top5drugs_dis %>% ggplot(aes(Product, tot_amt_reim, size = tot_amt_reim)) +
geom_point(aes(col = Product)) +
scale_size_continuous(range = c(1,20)) +
labs(title = "Top five drugs in country costing the taxpayers", x = "Drugs_name",
y = "Total cost (billions)") +
theme(title = black.bold.italic.16.text,
axis.title = black.bold.italic.16.text, axis.text = element_text(size = 10,
       face = 'bold')) +
scale_y_continuous(breaks = c(1000000000,3000000000,5000000000,7000000000),
labels = c(1.0,3.0,5.0,7.0)) + theme(legend.position = "none")
# TN

data_Drg_TN <- data_A %>% filter (State == 'TN') %>% 
  group_by(`Product`) %>% 
  summarize(tot_amt_reim = sum(`Total Amount Reimbursed`),
            tot_unt_reim = sum(`Units Reimbursed`),
            tot_nm_pres = sum(`Number of Prescriptions`),
            pop = mean(`Population`, na.rm = TRUE)) %>% 
  mutate(unit_cost = tot_amt_reim/tot_unt_reim) %>% 
  arrange(desc(tot_amt_reim))

top10drugs_TN <- head(data_Drg_TN, 10)
top5drugs_TN <- head(data_Drg_TN, 5)

top10drugs_TN_dis <- merge(top10drugs_TN,dis_prod, by = "Product" )
top5drugs_TN_dis <- merge(top5drugs_TN,dis_prod, by = "Product" )

top5drugs_TN_dis %>%
  ggplot(aes(Product, tot_amt_reim, size = tot_amt_reim)) +
  geom_point(aes(col = Product)) + scale_size_continuous(range = c(1,20)) +
  labs(title = "Top five drugs costing the taxpayers in TN", x = "Drugs_name",
  y = "Total_cost") +
  theme(title = black.bold.italic.16.text,
  axis.title = black.bold.italic.16.text, axis.text=element_text(size = 10,
       face = 'bold')) +
  scale_y_continuous(name = "Total cost (billions)",
  breaks = c(50000000,100000000,200000000),
  labels = c(0.05,0.1,0.2)) + theme(legend.position="none")


```


### **Diseases resposible for the high cost**

```{r, fig.height=8, fig.width=9.5}
# getting Total value with generic cost also
top10drugs_dis_generic <- top10drugs_dis %>% 
mutate (Tot_gen_cost = tot_unt_reim*gen_cost) %>% 
select ('Product', 'tot_amt_reim','Dis','Tot_gen_cost')
top10drugs_dis_generic_a <- top10drugs_dis_generic[c(1,2,4,8,10),]
# plotting data disease wise
top_diseases <- top10drugs_dis_generic %>%
  group_by(`Dis`) %>%
  summarize(total_cost = sum (`tot_amt_reim`),
  total_gen_cost = sum(`Tot_gen_cost`))

# plotting
    
    top_diseases %>% ggplot(aes(Dis, total_cost, size = total_cost)) +
     geom_point(aes(col = Dis)) +
     scale_size_continuous(range = c(1,20)) + 
     labs(title = "Diseases with high cost in country", x = "Disease_name",
          y = "Total cost (billions)") +
     theme(title = black.bold.italic.16.text,
    axis.title =   black.bold.italic.16.text,
    axis.text=element_text(size = 10,face = 'bold')) +
    scale_y_continuous(name = "Total cost (billions)",
    breaks = c(2000000000,4000000000,6000000000, 8000000000,10000000000),
    labels = c(2.0,4.0,6.0, 8.0, 10.0)) + theme(legend.position="none")
   
    # getting Total value with generic cost also
top10drugs_TN_dis_generic <- top10drugs_TN_dis %>% 
mutate (Tot_gen_cost = tot_unt_reim*gen_cost) %>% 
select ('Product', 'tot_amt_reim','Dis','Tot_gen_cost')

# plotting data disease wise
top_TN_diseases <- top10drugs_TN_dis_generic %>%
  group_by(`Dis`) %>%
  summarize(total_cost = sum (`tot_amt_reim`),
  total_gen_cost = sum(`Tot_gen_cost`))

# plotting
    
    top_TN_diseases %>% ggplot(aes(Dis, total_cost, size = total_cost)) +
     geom_point(aes(col = Dis)) +
     scale_size_continuous(range = c(1,20)) + 
     labs(title = "Disease with high cost in TN", x = "Disease_name",
          y = "Total cost (billions)") +
     theme(title = black.bold.italic.16.text,
    axis.title =   black.bold.italic.16.text,
    axis.text=element_text(size = 8,face = 'bold')) +
    scale_y_continuous(name = "Total cost (billions)",
    breaks = c(50000000,150000000,250000000,350000000),
    labels = c( 0.05,0.15,0.25,0.35)) + theme(legend.position="none")
   
```


### **Prescription vs Generic cost of top ten drugs **

```{r, fig.height=2, fig.width=4}

# changing the name of columns in top10drugs_dis_generic
   names(top10drugs_dis_generic_a)[names(top10drugs_dis_generic_a) ==
                                   "tot_amt_reim"] <- "Prescription"
   names(top10drugs_dis_generic_a)[names(top10drugs_dis_generic_a) ==
                                   "Tot_gen_cost"] <- "Generic"
  
  # making single column for pres_cost and gen_cost using gather function
comp_Prod <- gather(top10drugs_dis_generic_a, Drug, Cost, c("Prescription","Generic"))
   
 # Plotting  box plot showing difference
      p <- ggplot(comp_Prod, aes(x = Drug, y = Cost, fill = Drug)) +
     geom_boxplot() +
     labs(title = "Prescription vs Generic", x = "Drugs",
          y = "Total cost (billions)") +
     theme(title = black.bold.italic.16.text,
    axis.title =   black.bold.italic.16.text) +
     scale_y_continuous(breaks = c(2000000000,4000000000,6000000000,
                                8000000000,10000000000),
    labels = c(2.0,4.0,6.0, 8.0, 10.0)) + 
     theme(legend.position = "none")

    ggplotly(p)
   
  

```

### **What's the prediction for 2018 and 2019**

```{r, fig.height=1.8, fig.width=4}
 mod <- lm(total_cost ~ Year, data = Yearly_cost)
new_data <- data.frame(Year = c(2018, 2019))
exp_val <- broom::augment(mod, newdata = new_data)
xx <- ggplot(data = Yearly_cost, aes(y = total_cost, x = Year)) +
  geom_point() + geom_smooth(method = "lm") +
  geom_point(data = exp_val, aes(y = .fitted), size = 3, color = "red")+
  labs(title = "Medicaid cost over the years", x = "Year", y = "Total cost (billions)") +
theme(title = black.bold.italic.16.text, axis.title = black.bold.italic.12.text) +
scale_x_continuous(name="Year",breaks=c(2013,2014,2015,2016,2017,2018,2019),
labels = c(2013,2014,2015,2016,2017,2018,2019)) +
scale_y_continuous(name="Total cost (billions)", breaks = c(10000000000,
  20000000000,30000000000,40000000000,50000000000,60000000000,80000000000, 100000000000 ),labels = c(10,20,30,40,50, 60, 80,100))

ggplotly(xx)


```










